### 2022-11-10 chapter1.tex
Added non-breaking spaces to citations and references

\section*{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer \cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s \cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes \cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering \cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig. \ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results \cite{Thomas2020, Feng2020, Rhie2021}. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig. \ref{fig:graphical_overview}). We did not seek to reinvent the wheel, so the overall principle of clustering a graph of hits from homology searches is unchanged, though many details were modified to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, as modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were instead focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons \cite{Frankish2015, Dunne2018}. As a result, the structure and lengths of the protein sequences that result from these models can be highly heterogeneous, which poses many challenges for their alignment and subsequent analyses.

To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes. All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included \cite{D12GC2007}. However, all have been re-sequenced and re-annotated, many with long-read sequencing technology, which has resulted in substantial improvements in the contiguity of their assemblies. Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity \cite{Miller2018, Kim2021}. Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail \cite{Wiegmann2011, Obbard2012}.

@@

\section*{Results}
\subsection*{Input genomes}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete \cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype \cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficient in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase \cite{Gramates2022}. The complete list of genomes is given in Table \ref{stable:genomes}.

@@

\subsection*{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair \cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig. \ref{fig:components}). Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated. Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes. Others are possibly novel genes. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data creates many edges which form triangles but are otherwise poorly supported by the remaining genomes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig. \ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. \textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig. \ref{sfig:clusters}).

@@

\subsection*{Addition of paralogs to orthologous groups}
A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs. Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy. This is corrected by adding likely paralogs to the orthologous groups. Briefly, the protein sequences in each genome were searched against themselves. If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair. Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group. Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig. \ref{sfig:paralogs}).

@@

\subsection*{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence \cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table \ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction \cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns \cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig. \ref{fig:realignment}). Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.

@@

Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses. This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity \cite{Castresana2000, CapellaGutierrez2009}. This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold. Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns. Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig. \ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig. \ref{sfig:hmm_trim}).

@@

\section*{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteome and specific gene families in the \textit{Drosophila} genus. To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig. \ref{fig:trees}). Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.

@@

For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years \cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes. Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes \cite{Tatusov1997}. This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters. Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster. The optimal value of \textit{k} for a given set of genomes is unclear, but it likely depends on the desired trade-off between sensitivity and specificity. However, percolation theory or simulations may yield insights.

@@

The second major innovation in this work is its method for the refinement and trimming of alignments. In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile. This process usually corrects errors caused by the presence of long poorly supported segments. In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus. While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller \cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons \cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage} \cite{Nord2018}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.

@@

The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet. This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts. The transitions between amino acids could be parametrized with a published matrix such as LG \cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data. It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.

Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related \cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline \cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

>>>>

\section*{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results~\cite{Thomas2020, Feng2020, Rhie2021}. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}). We did not seek to reinvent the wheel, so the overall principle of clustering a graph of hits from homology searches is unchanged, though many details were modified to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, as modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were instead focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons~\cite{Frankish2015, Dunne2018}. As a result, the structure and lengths of the protein sequences that result from these models can be highly heterogeneous, which poses many challenges for their alignment and subsequent analyses.

To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes. All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included~\cite{D12GC2007}. However, all have been re-sequenced and re-annotated, many with long-read sequencing technology, which has resulted in substantial improvements in the contiguity of their assemblies. Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}. Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.

@@

\section*{Results}
\subsection*{Input genomes}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficient in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. The complete list of genomes is given in Table~\ref{stable:genomes}.

@@

\subsection*{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}). Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated. Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes. Others are possibly novel genes. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data creates many edges which form triangles but are otherwise poorly supported by the remaining genomes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. \textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Addition of paralogs to orthologous groups}
A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs. Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy. This is corrected by adding likely paralogs to the orthologous groups. Briefly, the protein sequences in each genome were searched against themselves. If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair. Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group. Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).

@@

\subsection*{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}). Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.

@@

Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses. This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}. This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold. Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns. Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig.~\ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.~\ref{sfig:hmm_trim}).

@@

\section*{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteome and specific gene families in the \textit{Drosophila} genus. To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}). Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.

@@

For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes. Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes~\cite{Tatusov1997}. This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters. Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster. The optimal value of \textit{k} for a given set of genomes is unclear, but it likely depends on the desired trade-off between sensitivity and specificity. However, percolation theory or simulations may yield insights.

@@

The second major innovation in this work is its method for the refinement and trimming of alignments. In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile. This process usually corrects errors caused by the presence of long poorly supported segments. In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus. While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons~\cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.

@@

The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet. This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts. The transitions between amino acids could be parametrized with a published matrix such as LG~\cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data. It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.

Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
